Quantum Hoare Types

### 2019-09-10 paper/paper.tex
Add rest of the poster

(2 pages)

@@

>>>>

The result will be a 2-qubit entangled state aka the first Bell state that will, on measurement, collapse into either \ket{00} or \ket{11} with equal probability.

\subsection{Desired Properties}
For our example, we would like to guarantee correctness of properties such as the following:
\begin{itemize}
	\item Before application of H gate, the qubit $q_0$ should have a deterministic integer value (on measurement)
	\item After application of H, $q_0$ should be in 50-50 probabilistic distribution of being in the two computational states
	\item After application of CNOT, the two qubits should be in a maximally entangled state
	\item Given a certain input to the whole circuit, we should get the corresponding Bell state
\end{itemize}

\section{Idea}
(2 pages)

\textit{The main idea of this paper is \ldots}

Recent work at Princeton as part of EPiQC\cite{DBLP:conf/oopsla/HuangM18,Huang:2019:SAV:3307650.3322213} has identified several classes of bugs in quantum programs and proposed approaches to tackle them. They proposed assertion checking using preconditions and post-conditions as one of the most effective techniques for most such classes of bugs.

As programming languages researchers, we think it will be more useful to encode such assertions into a static type system for both formal verification and aiding the programmers in writing correct programs from the start.

@@

\subsection{Notation and Background}
\begin{itemize}
	\item \textbf{Dependent Types}
	\[ \mathrm{inner\_product}: \Pi m: \mathrm{nat}.\mathrm{vector}(m) \times \mathrm{vector}(m) \rightarrow \mathrm{nat} \]

	\item \textbf{Hoare Triples} from \textbf{Floyd-Hoare Logic}
	\[ \{P\}\ C\ \{Q\} \]
	$P$ and $Q$ are pre and postcondition respectively on the program state, while $C$ is the command. For example:
	\[ \{\mathrm{True}\}\ x := 3\ \{x=3\} \]
	\item \textbf{Hoare Types} from \textbf{Hoare Type Theory} (\textbf{HTT}) that combine the strengths of dependent types and Hoare logic.
	\[ \{P\}\ \mathrm{res}:A\ \{Q\}\]
	for a stateful computation executed in a heap that satisfies precondition $P$ and returns a value of type $A$ in a heap that satisfies postcondition $Q$. For example:
	\[ \mathrm{alloc} : \forall \alpha . \Pi x : \alpha. \{\mathrm{emp}\}\ y : \mathrm{nat}\ \{ y \mapsto_{\alpha} x\}\]
\end{itemize}

\subsection{Quantum Hoare Types}
Quantum Hoare typing for the Hadamard and the CNOT operations will look something like:

\abovedisplayskip=0pt
\belowdisplayskip=12pt

\[ \mathbf{H}(q_0) : \{ \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\}\ r : \mathrm{unit}\ \{ \mathrm{q\_state}(q_0) \in \{\ket{+}, \ket{-}\} \} \]

\[ \mathbf{CNOT}(q_0, q_1) : \{ \mathrm{q\_state}(q_0) \in \{ \ket{+}, \ket{-} \} \wedge \mathrm{q\_state}(q_0) \in \{\ket{0}, \ket{1} \}\ r : \mathrm{unit}\ \{ \mathrm{entangled}(q_0, q_1) \}\]

Using sequential composition, type of the complete program will be:

\[ \mathbf{BellPair}(q_0, q_1) : \{ \mathrm{q\_state}(q_i) \in \{\ket{0}, \ket{1} \}\}\ r : \mathrm{unit}\ \{ \mathrm{entangled}(q_0, q_1) \} \]

where the predicates need to be encoded in terms of suitable dependent types. This is ongoing work.

\subsubsection{Challenges}
\begin{itemize}
	\item The type system for Hoare Type Theory is fairly complex. We are mechanizing the type system to handle design changes easily.
	\item We need to be able to reason about both decomposable quantum state and entangled state. This aspect may require incorporating ideas from Separation Logic.
\end{itemize}

\subsubsection{Benefits}
\begin{itemize}
	\item Potential to be a unified system for programming, specification and reasoning about quantum programs similar to HTT.
	\item Allows composition of verification in the same style as Hoare logic.
\end{itemize}

\hrulefill
